\section{Related Work}
\label{sec:related}

\paragraph{Cognitive maps in neural models.} The Tolman-Eichenbaum Machine
(TEM)~\citep{whittington2020tem} introduced transformer-adjacent models of
the hippocampal formation that factorize ``structure'' (position) from
``content'' (sensory observations). Its transformer variant,
TEM-T~\citep{whittington2022tem-t}, reformulated this as a transformer with
action-dependent position updates. MapFormer~\citep{rambaud2025mapformer}
is the most recent development in this line, replacing TEM-T's sequential
rotation updates with a parallel cumulative-sum Lie-algebra path integration.

\paragraph{Transformer position encodings.} RoPE~\citep{su2021roformer}
rotates queries/keys by angles that are a fixed function of token index;
CoPE~\citep{golovneva2024cope} gates rotations by content to compress
relative distance; TAPE~\citep{zhu2025tape} updates positional embeddings
after attention. None of these adapt rotations to an external action
signal. MapFormer's core contribution is precisely this adaptation. Per
the MapFormer paper's own Table 2, all three fail the 2D-navigation task
at a single layer (RoPE 0.16, TAPE 0.32, CoPE 0.74 IID) while
MapFormer-WM saturates at $\approx 1.00$.

\paragraph{Cognitive-map filtering frameworks.} The Clone-Structured
Cognitive Graph (CSCG;~\citealp{george2021cscg}) uses discrete latent
states --- ``clones'' --- to disambiguate aliased observations,
representing the Bayesian posterior over position as a categorical
distribution. This is a forward-model strategy: given a position, predict
the expected observation distribution. Classical navigation filters
(EKF, UKF, Invariant EKF) use inverse models: given an observation,
infer the implied position. Our analysis frames these as complementary
inductive biases for the correction mechanism; our main Level~1.5
variant is inverse-model-based, while our Predictive-Coding variant is
forward-model-based.

\paragraph{Invariant filtering on Lie groups.}
\citet{barrau2017inekf} introduced the Invariant Extended Kalman Filter,
showing that on matrix Lie groups with group-affine dynamics the error
admits autonomous (state-independent) dynamics, enabling bounded-error
guarantees under observability. \citet{markovic2017wrapping} showed that
on $\mathrm{SO}(2)$ the wrapped-innovation EKF is mathematically
equivalent to the full Lie-Group EKF, which we exploit for correctness
at minimal implementation cost.

\paragraph{Parallel Bayesian filtering.}
\citet{sarkka2021parallel} introduced parallelizable Bayesian filters
via associative scan, enabling $\mathcal{O}(\log T)$ depth for
linear-Gaussian state-space models. Subsequent work extended this to
nonlinear iterated filters. Mamba~\citep{gu2024mamba} brought such scans
to deep learning as a replacement for attention; our Level~2 design uses
the same associative-scan primitive on M\"obius $2\times 2$ matrices,
and Level~1.5 uses a simpler scalar-affine variant.

\paragraph{Predictive coding.} Rao \& Ballard (1999) formalized the
``prediction error as signal'' view: top-down predictions compared
against bottom-up signals give error signals that drive state updates.
Friston's Free-Energy Principle generalizes this. We implement a
predictive-coding variant of MapFormer that uses an MLP forward model
from $(\cos\theta, \sin\theta)$ to expected observation embedding, with
prediction error driving an additive state correction. We find it
complementary to InEKF's inverse-model approach --- stronger at aliased
aggregation, weaker at sharp landmark measurement.

\paragraph{State-space language models and the Mamba-Kalman connection.}
Selective SSMs \citep{gu2024mamba} use associative scans for
parallel-friendly recurrence. \citet{wang2025mamba-kalman} prove an
isomorphism: Mamba's selective SSM is equivalent to a time-varying
linear Kalman filter under an appropriate parameterization. Our Level~1.5
is a specialisation of this: a 1-D Mamba block with three additional
constraints --- (i) Kalman-ratio gain parameterization
$K_t = \Pi/(\Pi + R_t)$, (ii) Lie-group-correct wrapping of innovations,
(iii) decoupled path-integration + correction scans. We demonstrate
empirically that this specialisation is \emph{discoverable} from
gradient descent while generic selective SSMs are not, in the
cognitive-map regime.

\paragraph{Heteroscedastic Kalman filtering.} Learning per-observation
measurement noise is a well-studied idea in classical filtering (the
``innovation covariance estimation'' literature of the 1970s--80s) and
in modern learned-filter work such as KalmanNet (Revach et al., 2022).
Our Level~1.5 differs from these by coupling heteroscedastic $R_t$ with
fixed covariance $\Pi$, implemented via a single scalar associative
scan. This makes it architecturally lighter than KalmanNet while
retaining the key per-token gain adaptation.
