\section{Future Work}
\label{sec:future}

This paper focuses on a single correction mechanism (heteroscedastic
InEKF) in a single Lie group ($\mathrm{SO}(2)$) on a single task
(MapFormer's 2D torus navigation). Several natural extensions sit at
the intersection of classical robotics and modern deep learning, and
appear underexplored in the current literature.

\subsection{Higher-Dimensional Lie Groups}

MapFormer's $\mathrm{SO}(2)$ state is a natural starting point but
limits applications to planar navigation. Extensions:
\begin{itemize}
  \item \textbf{$\mathrm{SE}(2)$ (planar pose = position + heading):}
  direct generalization with no conceptual overhead; adds 2 translation
  components to the state. A natural fit for 2D robot navigation tasks
  (e.g., BabyAI, MiniGrid, ViZDoom).
  \item \textbf{$\mathrm{SO}(3)$ and $\mathrm{SE}(3)$ (3D orientation,
  full 6-DoF pose):} requires the 3D version of Barrau-Bonnabel's
  invariant EKF (well-understood in robotics). Would enable
  cognitive-map learning for drones, 3D navigation, and first-person
  VR/AR.
  \item \textbf{$\mathrm{Sim}(3)$ (pose with scale):} for monocular
  SLAM-style tasks where scale is uncertain. Used in ORB-SLAM,
  DROID-SLAM.
\end{itemize}
Our parallel-scan machinery extends directly: M\"obius matrices for
$\mathrm{SO}(2)$ generalize to Adjoint-weighted recurrences for
higher-dimensional groups following standard InEKF
derivations~\citep{sola2018micro}.

\subsection{IMU-Preintegration-Style Context Compression}

\citet{forster2017preintegration} compress 40 raw IMU samples into one
preintegrated $\mathrm{SE}(3)$ measurement for SLAM. The analogous
idea for long-context transformers: compress a chunk of
action/observation tokens into a single ``preintegrated'' position
update carrying a covariance.

Practically: replace the $\mathrm{cumsum}(\omega \cdot \Delta)$ over all
tokens with a multi-resolution representation that commits past chunks
into a single summary and keeps only recent tokens at full resolution.
Combined with Level~1.5's bounded-error property, this could extend
MapFormer's effective context dramatically without per-token attention.

\subsection{Equivariant Attention on Lie Groups}

RoPE is $\mathrm{SO}(2)$-equivariant by construction. Level~1.5
generalises via input-dependent rotations. But neither is
$\mathrm{SE}(n)$-equivariant --- they don't correctly handle the
coupling of translation and rotation. A natural extension:
$\mathrm{SE}(n)$-equivariant attention where Q/K rotate \emph{and}
translate under the group action. Recent work on equivariant
transformers~\citep{finzi2021practical} provides the machinery;
combining it with input-dependent actions (MapFormer's core innovation)
appears unexplored.

\subsection{Differentiable Pose-Graph SLAM Inside Neural Networks}

Theseus-AI~\citep{pineda2022theseus} provides a differentiable
$\mathrm{SE}(3)$ bundle adjustment module for PyTorch. MapFormer's
internal $\hat\theta$ state could be treated as pose estimates and
refined via a Theseus-style optimisation step at each layer or at
evaluation time. This would convert the ``cognitive map'' into a
literal pose-graph SLAM backend with learned perception.

\subsection{Connection to Biological Neural Data}
\label{sec:future:bio}

We tested two predictions empirically
(Section~\ref{sec:results:hippo}) and both came back \textbf{falsified}:
\begin{itemize}
  \item Hexagonal grid cells: Sargolini grid scores on hidden-state
  rate maps top out at 0.05--0.15 across all variants, well below the
  0.3 threshold considered grid-like.
  \item $R_t$ at landmark tokens: predicted ordering was
  \texttt{landmark} < \texttt{aliased} < \texttt{blank}; actual
  orderings deviate for both Level~1.5-WM and Level~1.5-EM.
\end{itemize}
The third test (Stensola $\sqrt{2}$-spaced $\omega$ modules) holds
approximately, but mostly inherited from the geometric initialisation
rather than discovered by training. The diagnostic explanation for the
absence of hexagonal cells is architectural --- see
Section~\ref{sec:future:grid}. Open directions:

\begin{itemize}
  \item Compare $R_t$ distributions to neural recordings from
  \citet{sun2024neural} or \citet{nieh2021geometry} directly. Even if
  Level~1.5 is not Bayesian-informativeness-optimal, the learned $R_t$
  may still match neural firing patterns under the right transformation.
  \item Train Level~1.5 with an explicit Bayesian-informativeness
  regulariser on $R_t$ to test whether the predicted ordering can be
  recovered.
  \item Investigate place-cell-like single-peak patterns at the
  \emph{attention} level rather than at hidden-state level --- attention
  rows project position onto a sparse subset of cells, which may exhibit
  place-cell selectivity even when hidden states do not show grid
  structure.
\end{itemize}

\subsection{Multi-Task / Instruction-Conditioned Navigation}

Current MapFormer evaluates ``predict the next observation at revisits.''
Real cognitive-map tasks involve goal-directed behavior: ``navigate to
the red door.'' This requires action prediction rather than observation
prediction, instruction conditioning (language or goal-state input), and
imitation learning or RL objectives. BabyAI is a natural next target.

\subsection{Scaling Study}

The MapFormer paper uses 1 layer, 2 heads, $d{=}128$ --- very small.
We replicate this scale but suspect the Level~1.5 advantages scale
favorably: more layers $\to$ more opportunities for $\hat\theta$
correction; more heads $\to$ per-head uncertainty could capture
multimodal hypotheses; larger $d$ $\to$ more expressive $R_t$ head. A
systematic scaling study at 2/4/8 layers, 4/8 heads,
$d{=}256/512/1024$ would establish whether the architectural advantage
persists at scale or shrinks.

\subsection{Multi-Environment Training and Richer Zero-Shot Transfer}

Our current pipeline trains each model on one fixed \texttt{obs\_map}
and evaluates zero-shot on held-out \texttt{obs\_map}s. TEM-style
multi-environment training would sample a fresh \texttt{obs\_map} per
batch during training; this tests whether residual gaps in the lm200
OOD regime close when the model has explicitly learned to generalise
across environments. Graph-topology generalisation (hexagonal grids,
family trees) is the deeper TEM-style test but requires more
substantial environment engineering.

\subsection{Real-World Robotics Deployment}

The Kalman regime of Level~1.5 --- bounded error, calibrated
uncertainty, sharp landmark handling --- is precisely what real
robotic navigation needs. We see a clear path from the current work to
practical impact: (1)~$\mathrm{SE}(3)$ generalisation of Level~1.5,
(2)~BabyAI or MiniGrid validation, (3)~sim-to-real on small mobile
robots (e.g., Clearpath Jackal), (4)~integration with existing SLAM
stacks via the Theseus-AI differentiable-optimisation bridge. This
would close the loop on the ``robotics Lie theory $\to$ deep learning''
transfer we advocate, with empirical validation on physical platforms.

\subsection{Comparison to Other Learned-Filter Frameworks}

KalmanNet (Revach et al.\ 2022) learns the Kalman gain while keeping the
rest classical. Level~1.5 learns both the gain (via $R_t$) and the
measurement model (via the $z$-head) but retains the Lie-group
structure. Understanding empirically where each framework wins (and
for which tasks) is a natural comparison study. Other frameworks for
comparison: particle filters with learned proposals, variational
Bayesian filters, Deep Kalman Filters.

\subsection{Unification with SSMs}

Level~1.5's scan machinery is algorithmically identical to Mamba's
selective-scan. This suggests deep connections between state-space
models (Mamba, S4), classical filtering (Kalman, EKF, UKF), and
equivariant architectures (Lie-group symmetry). A unified theoretical
framework covering all three would likely yield new architectures
combining the strengths of each. \citet{wang2025mamba-kalman} make
preliminary connections; extending this to the Lie-group setting with
explicit invariance guarantees is an open direction.

\subsection{MapFormer-Grid: An Architecture That Can Produce Hexagonal Cells}
\label{sec:future:grid}

Our hippocampal-correspondence analysis (Section~\ref{sec:future:bio})
found that current MapFormer variants do not produce hexagonal
grid-cell-like representations. The reason is structural: MapFormer
assigns one path-integrator block per $\omega$ frequency, but the
geometric optimum for 2D spatial tiling --- three sinusoidal waves at
the \emph{same} frequency oriented at $60^\circ$ --- requires multiple
blocks per scale. With one wave per $\omega$, hexagonal interference
patterns are mathematically inaccessible regardless of how the model is
trained.

\paragraph{Proposed architectural change.} Replace the single block per
$\omega$ with a \emph{module} of $n_\mathrm{orientations}$ blocks at
the same $\omega$ with orientations $\{0^\circ, 60^\circ, 120^\circ\}$
in 2D action space:
\begin{itemize}
  \item \texttt{action\_to\_lie} outputs a 2D vector
  $(\Delta_x, \Delta_y)$ per module instead of a scalar per block;
  \item for each block in module $m$ at orientation $\theta_o$:
  $\Delta_{m,o} = \cos(\theta_o) \cdot \Delta_x + \sin(\theta_o) \cdot \Delta_y$;
  \item the cumulative angle is
  $\theta_{m,o,t} = \omega_m \cdot \mathrm{cumsum}(\Delta_{m,o})$;
  \item at the hidden-state level,
  $\cos\theta_{m,0} + \cos\theta_{m,60} + \cos\theta_{m,120}$
  produces hexagonal interference at scale $\omega_m$.
\end{itemize}

\paragraph{Trade-offs.} More parameters per module, fewer distinct
$\omega$ values. Path integration remains parallelisable via cumsum;
$\mathcal{O}(\log T)$ preserved. Level~1.5 correction extends naturally
to the new state. A scaffold class \texttt{MapFormerWM\_Grid} exists in
the repository; multi-seed training and hippocampal-correspondence
evaluation are the obvious next step.

\paragraph{Predicted outcomes.}
(1) Hexagonal patterns emerge with grid scores $> 0.3$ at multiple
scales, matching \citet{sargolini2006conjunctive};
(2) multi-scale modular organisation matching
\citet{stensola2012modules} $\sqrt{2}$-ratio spacing;
(3) cognitive-map task performance preserved or improved, since
hexagonal codes are theoretically optimal for 2D position under
bandwidth constraints.

If validated, this would close the gap between MapFormer's mathematical
structure and biological grid-cell organisation, providing the first
transformer-based architecture that \emph{necessarily} produces
grid-cell representations from first principles. Falsification would be
a significant negative result indicating the gradient landscape itself
disfavours hexagonal solutions even when architecturally accessible.

\subsection{Robust Level 1.5-EM Initialization}
\label{sec:future:em-init}

Level~1.5-EM currently requires a safe initialisation
($\log R_\mathrm{init\_bias} = 3.0$) because EM's Hadamard product
attention provides no fallback gradient path when the position branch
is corrupted by random $\hat\theta$ corrections at init. Our experiment
with a more aggressive safe init (bias $= 5.0$) showed that no single
bias is optimal for all seeds --- the lm200 variance is intrinsic to
seed-specific landscape geometry.

A more principled fix could address this. Possibilities: (i) a learnable
gate that replaces the correction at init with the identity and unfreezes
during training (adapter-style); (ii) explicit warmup scheduling of the
Kalman gain $K_t$ (e.g., from 0 to its learned value over the first
10\% of training); (iii) a different attention structure that does not
have the $A_P$ destruction failure mode but retains EM's
Hadamard-product expressivity. Closing this gap would make
Level~1.5-EM as robust and reliable as Level~1.5-WM, which currently is
the better-behaved of the two.
